# Prepare data

## Load gut data

Load the original data files outputted by the bioinformatic pipeline.

### Sample metadata

```{r load_sample_metadata_gut, warning=FALSE, comments="", message=FALSE, eval=FALSE}
sample_metadata <- read_csv2("data/squirrels_metadata.csv") %>% 
  arrange(sample)
```

### Genome metadata

```{r load_genome_metadata, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_metadata <- read_csv("data/squirrels_mag_info.csv") %>% 
  rename(length=mag_size) %>% 
  arrange(genome)
```

### Sequencing reads

```{r load_read_counts, warning=FALSE, comments="", message=FALSE, eval=FALSE}
read_counts <- read_csv2("data/squirrels_counts.csv") %>%
    rename(genome=1) %>%
    arrange(match(genome, genome_metadata$genome)) %>%
    select(c("genome",sample_metadata$sample))
```

```{r load_genome_coverage, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_coverage <- read_csv2("data/squirrels_coverage.csv") %>%
    rename(genome=1) %>%
    arrange(match(genome, genome_metadata$genome)) %>%
    select(c("genome",sample_metadata$sample))
```

### Genome tree

```{r load_genome_tree_gut, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_tree <- read_tree("data/squirrels.tree")
genome_tree$tip.label <- str_replace_all(genome_tree$tip.label,"'", "") #remove single quotes in MAG names
genome_tree <- keep.tip(genome_tree, tip=genome_metadata$genome) # keep only MAG tips
```


### Genome annotations

```{r load_genome_annotations, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_annotations <- read_tsv("data/genome_annotations.tsv.xz") %>%
    rename(gene=1, genome=2, contig=3)
```

## Create working objects

Transform the original data files into working objects for downstream analyses.

### Filter reads by coverage

```{r filter_coverage, warning=FALSE, comments="", message=FALSE, eval=FALSE}
min_coverage=0.3
read_counts_filt <- genome_coverage %>%
  mutate(across(where(is.numeric), ~ ifelse(. > min_coverage, 1, 0))) %>%
  mutate(across(-1, ~ as.numeric(.) * as.numeric(read_counts[[cur_column()]])))
```


### Transform reads into genome counts

```{r calculate_genome_counts_unfiltered, warning=FALSE, comments="", message=FALSE, eval=FALSE}
readlength=150
genome_counts <- read_counts %>%
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / readlength) ))
```

```{r calculate_genome_counts_filtered, warning=FALSE, comments="", message=FALSE, eval=FALSE}
readlength=150
genome_counts_filt <- read_counts_filt %>%
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / readlength) ))
```

### Distill annotations into GIFTs

```{r distill_annotations_gut, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_gifts <- distill(genome_annotations,GIFT_db2,genomecol=2,annotcol=c(9,10,19), verbosity=F)
```

## Prepare color scheme

[AlberdiLab](www.alberdilab.dk) projects use unified color schemes developed for the [Earth Hologenome Initiative](www.earthhologenome.org), to facilitate figure interpretation.

```{r get_ehi_colors_gut, warning=FALSE, comments="", message=FALSE, eval=FALSE}
ehi_phylum_colors <- read.table("data/ehi_phylum_colors.tsv",sep="\t",header=T,comment.char = "")

phylum_colors <- ehi_phylum_colors %>%
  mutate(colors2 = if_else(phylum=="p__Bacillota", '#074759',
          if_else(phylum=="p__Bacillota_A", '#0c80a2',
          if_else(phylum=="p__Bacillota_B", '#10acd9',
          if_else(phylum=="p__Bacillota_C", '#7ddaf5',
          if_else(phylum=="p__Actinomycetota", '#C7B3E0',
          if_else(phylum=="p__Pseudomonadota", '#f5aa74',
          if_else(phylum=="p__Thermoplasmatota", '#f07e2a',
          if_else(phylum=="p__Verrucomicrobiota", '#77638A',
          if_else(phylum=="p__Cyanobacteriota", '#94cc84',
          if_else(phylum=="p__Elusimicrobiota", '#226d1b', colors
                  )))))))))))
#9E8DAD
#08D1D1
#ccffbc
#5c9c4f
#08a26b

ehi_colors <- phylum_colors %>% pull(colors)
names(ehi_colors) <- phylum_colors$phylum

custom_colors <- phylum_colors %>% pull(colors2)
names(custom_colors) <- phylum_colors$phylum

gift_colors <- read.table("data/gift_colors.tsv",sep="\t",header=T,comment.char = "")

squirrel_colors <- c("#999999", "#cc3333")
names(squirrel_colors) <- c("Sciurus carolinensis", "Sciurus vulgaris")

season_colors <- c("#76b183","#d98a41","#6b7398") #e5bd5b
names(season_colors) <- c("spring-summer", "autumn", "winter")

area_colors <- c("#3d6a4b","#94a766","#c1cca7")
names(area_colors) <- c("rural", "suburban", "urban")

sex_colors <- c("turquoise3", "indianred2")
names(sex_colors) <- c("M", "F")
```

```{r get_colors_gut, warning=FALSE, comments="", message=FALSE, eval=FALSE}
altitude_colors<- c("#429ef5","#a8c3f0","#f56c42","#f29c80")
treatment_colors <- c("#429ef5","#f56042")
```

#prepare a phyloseq object
phylo_samples <- sample_metadata %>% 
  column_to_rownames("sample") %>% 
  sample_data() #convert to phyloseq sample_data object
phylo_genome <- genome_counts_filt %>% 
  column_to_rownames("genome") %>% 
  otu_table(., taxa_are_rows = TRUE)
phylo_taxonomy <- genome_metadata %>%
  column_to_rownames("genome") %>% 
  as.matrix() %>% 
  tax_table() #convert to phyloseq tax_table object
phylo_tree <- phy_tree(genome_tree) 

physeq_genome <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples,phylo_tree)
physeq_genome_clr <- microbiome::transform(physeq_genome, 'clr')

# Wrap working objects

All working objects are wrapped into a single Rdata object to facilitate downstream usage.

```{r wrap_working_objects_gut, warning=FALSE, comments="", message=FALSE, eval=FALSE}
save(sample_metadata, 
     genome_metadata, 
     read_counts, 
     genome_counts, 
     genome_counts_filt, 
     genome_tree,
     physeq_genome,
     phylum_colors,
     file = "data/data.Rdata")
```
git remote set-url origin https://<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>@github.com/alberdilab/squirrels_functional_diversity.git

#sequencing reads statisitcs
```{r}
sample_metadata %>% 
    summarise(Total=sum(reads_post_fastp * 150 / 1000000000) %>% round(2), 
              mean=mean(reads_post_fastp * 150 / 1000000000) %>% round(2),
              sd=sd(reads_post_fastp * 150 / 1000000000) %>% round(2)) %>%
    unite("Average",mean, sd, sep = " ± ", remove = TRUE) %>%
    tt()
```

#DNA fractions
```{r}
sequence_fractions <- read_counts %>%
  pivot_longer(-genome, names_to = "sample", values_to = "value") %>%
  group_by(sample) %>%
  summarise(mags = sum(value)) %>%
    left_join(sample_metadata, by = join_by(sample == sample)) %>%
    select(sample,mags,metagenomic_bases,host_bases,bases_lost_fastp_percent) %>%
    mutate(mags_bases = mags*146) %>%
    mutate(lowqual_bases = ((metagenomic_bases+host_bases)/(1-as.numeric(bases_lost_fastp_percent)))-(metagenomic_bases+host_bases)) %>%
    mutate(unmapped_bases = metagenomic_bases - mags_bases) %>%
    mutate(unmapped_bases = ifelse(unmapped_bases < 0, 0, unmapped_bases)) %>%
    select(sample, lowqual_bases, host_bases, unmapped_bases, mags_bases)

sequence_fractions %>%
  mutate_at(vars(-sample), ~./1000000000) %>%
  rename("Sample"=1, "Low quality"=2, "Mapped to host"=3, "Unmapped"=4, "Mapped to MAGs"=5) %>%
  tt()
```

#sequence fractions
```{r}
sequence_fractions %>%
  mutate_at(vars(-sample), ~./1000000000) %>%
  rename("Sample"=1, "Low quality"=2, "Mapped to host"=3, "Unmapped"=4, "Mapped to MAGs"=5) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE)) %>%
  tt()
```
```{r}
sequence_fractions %>%
    pivot_longer(!sample, names_to = "fraction", values_to = "value") %>%
    mutate(value = value / 1000000000) %>%
    mutate(fraction = factor(fraction, levels = c("lowqual_bases","host_bases","unmapped_bases","mags_bases"))) %>%
  
    ggplot(., aes(x = sample, y = value, fill=fraction)) +
        geom_bar(position="stack", stat = "identity") +
      scale_fill_manual(name="Sequence type",
                    breaks=c("lowqual_bases","host_bases","unmapped_bases","mags_bases"),
                    labels=c("Low quality","Mapped to host","Unmapped","Mapped to MAGs"),
                    values=c("#CCCCCC", "#bcdee1", "#d8b8a3","#93655c"))+
        labs(x = "Samples", y = "Amount of data (GB)") +
        theme_classic() +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size=6),legend.position = "bottom")
```
#recovered microbial fraction
```{r data_estimations_plot, message=FALSE, warning=FALSE, fig.height=6, fig.width=10, fig.fullwidth=TRUE}}
singlem_table <- sequence_fractions %>%
	mutate(mags_proportion = round((mags_bases / (mags_bases + unmapped_bases))*100,2)) %>%
  left_join(sample_metadata, by = join_by(sample == sample))  %>%
	mutate(singlem_proportion = round(as.numeric(sample_metadata$singlem_fraction)*100,2)) %>%
	select(sample,mags_proportion,singlem_proportion) %>%
	mutate(mags_proportion = ifelse(singlem_proportion == 0, 0, mags_proportion)) %>% #convert zeros to NA
	mutate(singlem_proportion = ifelse(singlem_proportion == 0, NA, singlem_proportion)) %>% #convert zeros to NA
	mutate(singlem_proportion = ifelse(singlem_proportion < mags_proportion, NA, singlem_proportion)) %>% #if singlem is smaller, then NA, to simplify plot
	mutate(singlem_proportion = ifelse(singlem_proportion > 100, 100, singlem_proportion)) #simplify

singlem_table %>%
	pivot_longer(!sample, names_to = "proportion", values_to = "value") %>%
	mutate(proportion = factor(proportion, levels = c("mags_proportion","singlem_proportion"))) %>%
	ggplot(., aes(x = value, y = sample, color=proportion)) +
			geom_line(aes(group = sample), color = "#f8a538") +
			geom_point() +
      scale_color_manual(name="Proportion",
                    breaks=c("mags_proportion","singlem_proportion"),
                    labels=c("Recovered","Estimated"),
                    values=c("#52e1e8", "#876b53"))+
			theme_classic() +
			labs(y = "Samples", x = "Prokaryotic fraction (%)") +
	    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size=6),legend.position = "right")
```

```{r}
# Export difference b/w mags and singlem proportions to be used later in script 05-diversity_models
singlem <- sequence_fractions %>%
  mutate(mags_proportion = round((mags_bases / (mags_bases + unmapped_bases))*100,2)) %>%
  left_join(sample_metadata, by = join_by(sample == sample))  %>%
  mutate(singlem_proportion = round(as.numeric(singlem_fraction)*100,2)) %>%
  mutate(mags_singlem = mags_proportion/singlem_proportion) %>%
  mutate(est_mapp = ifelse(mags_singlem >= 1, 1, mags_singlem)) %>%
  select(sample,mags_singlem,est_mapp)

write.table(singlem, file = "data/singlem.csv", row.names = FALSE, dec = ".", sep = ";", 
              quote = FALSE)
```

```{r}
damr <- singlem_table %>%
  mutate(damr=ifelse(is.na(singlem_proportion),100,mags_proportion/singlem_proportion*100)) %>%
  select(sample,damr)

damr %>% 
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  summarise(mean=mean(damr),sd=sd(damr)) %>% 
  tt()
```

```{r}
damr %>% 
  tt()
```

```{r}
damr %>% 
  filter(damr>95) %>%
  nrow()
```

```{r}
damr %>% 
  summarise(mean=mean(damr),sd=sd(damr)) %>% 
  tt()
```

#mag catalouge
```{r}

genome_counts_filt <- genome_counts %>% 
    select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>% 
  semi_join(., genome_counts_filt, by = "genome") %>% 
  arrange(match(genome,genome_counts_filt$genome))

genome_tree <- keep.tip(genome_tree, tip=genome_metadata$genome) # keep only MAG tips

#load("data/genome_gifts.Rdata")
```


#genome phylogeny
phylum_colors <- as.data.frame(phylum_colors)
```{r}
# Which phylum the MAG belongs to
phyla <- phylum_colors %>%
  right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
  arrange(match(genome, genome_tree$tip.label)) %>%
  select(phylum, colors2) %>%
  unique()

# What is the genome size of the MAG in MBs (megabases)
lengths <- genome_metadata %>%
  select(c(genome,length)) %>%
  mutate(length=round(length/1000000,2))

# What is the completeness of the MAG
mag_completeness <- genome_metadata %>%
  select(c(genome,completeness)) %>%
  as.data.frame() %>%
  remove_rownames() %>%
  column_to_rownames(var = "genome")


# Generate the phylum color heatmap
phylum_heatmap <- phylum_colors %>%
  right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
  arrange(match(genome, genome_tree$tip.label)) %>%
  select(genome,phylum) %>%
  mutate(phylum = factor(phylum, levels = unique(phylum))) %>%
  column_to_rownames(var = "genome")

# Create baseline circular genome tree
circular_tree <- force.ultrametric(genome_tree,method="extend") %>%
  ggtree(., layout = 'circular', size = 0.1, angle=45) +
  xlim(-1, NA)

# Add phylum colors ring
circular_tree <- gheatmap(circular_tree, phylum_heatmap, offset=0.05, width=0.3, colnames=FALSE) +
  scale_fill_manual(values=custom_colors, name="Phylum") +
  #geom_tiplab2(size=1, hjust=-0.1) +
  theme(legend.position = "right", plot.margin = margin(0, 0, 0, 0), panel.margin = margin(0, 0, 0, 0))

# Flush color scale to enable a new color scheme in the next ring
circular_tree <- circular_tree + new_scale_fill()

# Add completeness ring
circular_tree <-  circular_tree +
  #new_scale_fill() +
  scale_fill_gradient(low = "#d1f4ba", high = "#f4baba", name="Contamination") +
  geom_fruit(
    data=genome_metadata,
    geom=geom_bar,
    mapping = aes(x=completeness, y=genome, fill=contamination),
    offset = 0.55,
    orientation="y",
    stat="identity")

# Add genome-size ring
circular_tree <-  circular_tree +
  new_scale_fill() +
  scale_fill_manual(values = "#cccccc") +
  geom_fruit(
    data=lengths,
    geom=geom_bar,
    mapping = aes(x=length, y=genome),
    offset = 0.05,
    orientation="y",
    stat="identity")


#Plot circular tree
circular_tree

```

#taxonomy overview
```{r}
tax_mag <-genome_metadata %>% 
  group_by(phylum) %>%
  summarise(mag_n=n()) 

tax_mag %>%
  mutate(percetage_mag=round(mag_n*100/sum(mag_n), 2)) %>% 
  arrange(-percetage_mag) %>%
  tt()
```

#MAG size
```{r}
genome_metadata <-genome_metadata %>% 
  mutate(corrected_size=100*length/completeness)
```
```{r}
#MAG average size
genome_metadata %>% 
  summarise(Average_corrected_size=mean(corrected_size)) %>% 
  tt()
```
```{r}
#minimum MAG size
genome_metadata %>% 
  filter(corrected_size==min(corrected_size)) %>% 
  tt()
```
```{r}
#maximum MAG size
genome_metadata %>% 
  filter(corrected_size==max(corrected_size)) %>% 
  tt()
```
```{r}
genome_metadata %>% 
  arrange(corrected_size) %>% 
  paged_table()
```
```{r}
#average size and completeness
genome_metadata %>% 
  group_by(phylum) %>%
  summarise(average_size=mean(corrected_size),sd_size=sd(corrected_size),average_comp=mean(completeness),sd_comp=sd(completeness)) %>%
  mutate(Size=str_c(round(average_size,2),"±",round(sd_size,2)),
         Completeness=str_c(round(average_comp,2),"±",round(sd_comp,2))) %>% 
  arrange(-average_size) %>% 
  select(phylum, Size, Completeness) %>% 
  tt()
```

#Genome quality
```{r}
genome_metadata %>% 
    summarise(completeness_mean=mean(completeness) %>% round(2) %>% as.character(), 
              completeness_sd=sd(completeness) %>% round(2) %>% as.character(), 
              contamination_mean=mean(contamination) %>% round(2), 
              contamination_sd=sd(contamination) %>% round(2)) %>%
    unite("Completeness",completeness_mean, completeness_sd, sep = " ± ", remove = TRUE) %>%
    unite("Contamination",contamination_mean, contamination_sd, sep = " ± ", remove = TRUE) %>%
    tt()
```

```{r}
#create input table from original genome table
genome_details <- genome_metadata %>%
  select(c(genome,domain,phylum,completeness,contamination,length)) %>%
  mutate(length=round(length/1000000,2)) %>% #change length to MBs
  rename(comp=completeness,cont=contamination,size=length) %>% #rename columns
  remove_rownames() %>%
  arrange(match(genome, rev(genome_tree$tip.label))) #sort MAGs according to phylogenetic tree

#generate genome quality biplot
genome_stats_biplot <- genome_details %>%
  ggplot(aes(x=comp,y=cont,size=size,color=phylum)) +
  geom_point(alpha=0.7) +
  ylim(c(10,0)) +
  scale_color_manual(values=custom_colors) +
  labs(y= "Contamination", x = "Completeness") +
  theme_classic() +
  theme(legend.position = "none")

#generate contamination boxplot
genome_stats_cont <- genome_details %>%
  ggplot(aes(y=cont)) +
  ylim(c(10,0)) +
  geom_boxplot(colour = "#999999", fill="#cccccc") +
  theme_void() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        plot.margin = unit(c(0, 0, 0.40, 0),"inches")) #add bottom-margin (top, right, bottom, left)

#generate completeness boxplot
genome_stats_comp <-genome_details %>%
  ggplot(aes(x=comp)) +
  xlim(c(50,100)) +
  geom_boxplot(colour = "#999999", fill="#cccccc") +
  theme_void() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        plot.margin = unit(c(0, 0, 0, 0.50),"inches")) #add left-margin (top, right, bottom, left)

#create composite figure
grid.arrange(grobs = list(genome_stats_comp,genome_stats_biplot,genome_stats_cont),
             layout_matrix = rbind(c(1,1,1,1,1,1,1,1,1,1,1,4),
                                   c(2,2,2,2,2,2,2,2,2,2,2,3),
                                   c(2,2,2,2,2,2,2,2,2,2,2,3),
                                   c(2,2,2,2,2,2,2,2,2,2,2,3),
                                   c(2,2,2,2,2,2,2,2,2,2,2,3),
                                   c(2,2,2,2,2,2,2,2,2,2,2,3),
                                   c(2,2,2,2,2,2,2,2,2,2,2,3),
                                   c(2,2,2,2,2,2,2,2,2,2,2,3),
                                   c(2,2,2,2,2,2,2,2,2,2,2,3),
                                   c(2,2,2,2,2,2,2,2,2,2,2,3),
                                   c(2,2,2,2,2,2,2,2,2,2,2,3),
                                   c(2,2,2,2,2,2,2,2,2,2,2,3)))
```

#functional overview
```{r}
genome_annotations <- read_tsv("data/genome_annotations.tsv.xz") %>%
    rename(gene=1, genome=2, contig=3)

# Predicted genes
genome_annotations %>%
  nrow()
# Some annotation
genome_annotations %>%
  filter(if_any(c(kegg_id, pfam_hits, cazy_hits), ~ !is.na(.))) %>% 
  nrow()
# KEGG annotation
genome_annotations %>%
  filter(!is.na(kegg_id)) %>% 
  nrow()
```

```{r func_attributes, fig.dim = c(10, 8), fig.fullwidth=TRUE, warning=FALSE, message=FALSE} 


#Generate a basal utrametric tree for the sake of visualisation
gift_tree <- force.ultrametric(genome_tree,method="extend") %>%
   ggtree(., size = 0.3)

#Add phylum colors next to the tree tips
gift_tree <- gheatmap(gift_tree, phylum_heatmap, offset=0, width=0.1, colnames=FALSE, color=NA) +
   scale_fill_manual(values=custom_colors) +
    labs(fill="Phylum")

#Reset fill scale to use a different colour profile in the heatmap
gift_tree <- gift_tree + new_scale_fill()


#Add functions heatmap
gift_tree <- gheatmap(gift_tree, genome_gifts, offset=0.5, width=3.5, colnames=FALSE, color=NA) +
  vexpand(.08) +
  coord_cartesian(clip = "off") +
  scale_fill_gradient(low = "#f4f4f4", high = "steelblue", na.value="white")+
  labs(fill="GIFT")

#Reset fill scale to use a different colour profile in the heatmap
gift_tree <- gift_tree + new_scale_fill()

# Add completeness barplots
gift_tree <- gift_tree +
            geom_fruit(data=genome_metadata,
            geom=geom_bar,
            #grid.params=list(axis="x", text.size=2, nbreak = 1),
            axis.params=list(vline=TRUE),
            mapping = aes(x=length, y=genome, fill=completeness),
                 offset = 3.8,
                 orientation="y",
                 stat="identity") +
            scale_fill_gradient(low = "#cf8888", high = "#a2cc87") +
            labs(fill="Genome\ncompleteness")

#Plot combined tree + heatmap
gift_tree +
  theme(legend.position='right')

```
#functional ordination of MAGS
```{r, fig.width=10}


# Generate the tSNE ordination
tSNE_func2 <- Rtsne(X=genome_gifts, dims = 2, check_duplicates = FALSE)


# Plot the ordination
tSNE_func2$Y %>%
  as.data.frame() %>%
  mutate(genome=rownames(genome_gifts)) %>%
  inner_join(genome_metadata, by="genome") %>%
  rename(tSNE1="V1", tSNE2="V2") %>%
  select(genome,phylum,tSNE1,tSNE2, completeness) %>%
  ggplot(aes(x = tSNE1, y = tSNE2, color = phylum, size=completeness))+
  geom_point(shape=16, alpha=0.6) +
  scale_color_manual(values=custom_colors) +
  theme_minimal() +
  theme(legend.position = "right") + 
  guides(colour = guide_legend(override.aes = list(size=5, alpha=0.9)))

```

#Community composition
```{r data_preparation03, warning=FALSE}

genome_counts_log <- genome_counts %>% 
    column_to_rownames(var="genome") %>%
    mutate_all(~log10(.+1)) #fixed: mutate_at(vars(), ~log10(.+1))) was not working

genome_counts_pivot <- genome_counts %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS normalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) 
# %>% #append taxonomy
  # mutate(phylum = fct_relevel(phylum, rev(ehi_phylum_colors$phylum))) #sort phyla by taxonomy

genome_counts_by_host <- sample_metadata %>%
  select("sample","species","area_type", "development") %>%
  rename(host_sp=species) %>%
  left_join(genome_counts_pivot,., by=join_by("sample" == "sample")) #%>%

  
# Which host species each genome can be found in
genomes_by_species <- genome_counts_by_host %>%
  filter(count>0) %>%
  group_by(genome) %>%
  mutate(host = if_else(all(host_sp == "Sciurus vulgaris"), "only red",
                        if_else(all(host_sp == "Sciurus carolinensis"), "only grey", "both"))) %>%
  select(genome, host) %>%
  distinct(genome, .keep_all = TRUE) %>%
  left_join(.,genome_metadata, by='genome')

genomes_by_species$host <-factor(genomes_by_species$host, levels = c("both", "only red", "only grey"))
```

#overview by host species
```{r mags_by_species_tree, fig.height=8, fig.width=10, fig.fullwidth=TRUE, warning=FALSE, message=FALSE}

# Generate the phylum color heatmap
phylum_heatmap <- genome_metadata %>%
  arrange(match(genome, genome_tree$tip.label)) %>%
  select(genome,phylum) %>%
  mutate(phylum = factor(phylum, levels = unique(phylum))) %>%
  column_to_rownames(var = "genome")

# Create baseline circular genome tree
circular_tree <- force.ultrametric(genome_tree,method="extend") %>%
  ggtree(., layout = 'circular', size = 0.1, angle=45) +
  xlim(-1, NA)

# Add phylum colors ring
circular_tree <- gheatmap(circular_tree, phylum_heatmap, offset=0.0, width=0.3, colnames=FALSE) +
  scale_fill_manual(values=custom_colors, name="Phylum") +
  #geom_tiplab2(size=1, hjust=-0.1) +
  theme(legend.position = "right", plot.margin = margin(0, 0, 0, 0), panel.margin = margin(0, 0, 0, 0))


#Flush color scale to enable a new color scheme in the next ring
circular_tree <- circular_tree + new_scale_fill()

# Add host ring
circular_tree <-  circular_tree +
  new_scale_fill() +
  scale_fill_manual(values = c("black", "#ed2939", "#92a0ad"), name="Host\nspecies") + #"#cc3333", "#999999"
  geom_fruit(
    data=genomes_by_species,
    geom=geom_tile,
    mapping = aes(y=genome, fill=host),
    offset = 0.55,
    width=0.3)

#Plot circular tree
circular_tree


#MAGs by host species
genomes_by_species %>%
  dplyr::group_by(host) %>%
  summarise(n=length(host),
            percentage=(length(host)/1687)*100) %>%
  tt()
```




