# Prepare data

## Load gut data

Load the original data files outputted by the bioinformatic pipeline.

### Sample metadata

```{r load_sample_metadata_gut, warning=FALSE, comments="", message=FALSE, eval=FALSE}
sample_metadata <- read_csv2("data/squirrels_metadata.csv") %>% 
  arrange(sample)
```

### Genome metadata

```{r load_genome_metadata, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_metadata <- read_csv("data/squirrels_mag_info.csv") %>% 
  rename(length=mag_size) %>% 
  arrange(genome)
```

### Sequencing reads

```{r load_read_counts, warning=FALSE, comments="", message=FALSE, eval=FALSE}
read_counts <- read_csv2("data/squirrels_counts.csv") %>%
    rename(genome=1) %>%
    arrange(match(genome, genome_metadata$genome)) %>%
    select(c("genome",sample_metadata$sample))
```

```{r load_genome_coverage, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_coverage <- read_csv2("data/squirrels_coverage.csv") %>%
    rename(genome=1) %>%
    arrange(match(genome, genome_metadata$genome)) %>%
    select(c("genome",sample_metadata$sample))
```

### Genome tree

```{r load_genome_tree_gut, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_tree <- read_tree("data/squirrels.tree")
genome_tree$tip.label <- str_replace_all(genome_tree$tip.label,"'", "") #remove single quotes in MAG names
genome_tree <- keep.tip(genome_tree, tip=genome_metadata$genome) # keep only MAG tips
```


### Genome annotations

```{r load_genome_annotations, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_annotations <- read_tsv("data/genome_annotations.tsv.xz") %>%
    rename(gene=1, genome=2, contig=3)
```

## Create working objects

Transform the original data files into working objects for downstream analyses.

### Filter reads by coverage

```{r filter_coverage, warning=FALSE, comments="", message=FALSE, eval=FALSE}
min_coverage=0.3
read_counts_filt <- genome_coverage %>%
  mutate(across(where(is.numeric), ~ ifelse(. > min_coverage, 1, 0))) %>%
  mutate(across(-1, ~ as.numeric(.) * as.numeric(read_counts[[cur_column()]])))
```


### Transform reads into genome counts

```{r calculate_genome_counts_unfiltered, warning=FALSE, comments="", message=FALSE, eval=FALSE}
readlength=150
genome_counts <- read_counts %>%
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / readlength) ))
```

```{r calculate_genome_counts_filtered, warning=FALSE, comments="", message=FALSE, eval=FALSE}
readlength=150
genome_counts_filt <- read_counts_filt %>%
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / readlength) ))
```

### Distill annotations into GIFTs

```{r distill_annotations_gut, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_gifts <- distill(genome_annotations,GIFT_db,genomecol=2,annotcol=c(9,10,19), verbosity=F)
```

## Prepare color scheme

[AlberdiLab](www.alberdilab.dk) projects use unified color schemes developed for the [Earth Hologenome Initiative](www.earthhologenome.org), to facilitate figure interpretation.

```{r get_ehi_colors_gut, warning=FALSE, comments="", message=FALSE, eval=FALSE}
phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(phylum, colors) %>%
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name=phylum)
```

```{r get_colors_gut, warning=FALSE, comments="", message=FALSE, eval=FALSE}
altitude_colors<- c("#429ef5","#a8c3f0","#f56c42","#f29c80")
treatment_colors <- c("#429ef5","#f56042")
```

#prepare a phyloseq object
phylo_samples <- sample_metadata %>% 
  column_to_rownames("sample") %>% 
  sample_data() #convert to phyloseq sample_data object
phylo_genome <- genome_counts_filt %>% 
  column_to_rownames("genome") %>% 
  otu_table(., taxa_are_rows = TRUE)
phylo_taxonomy <- genome_metadata %>%
  column_to_rownames("genome") %>% 
  as.matrix() %>% 
  tax_table() #convert to phyloseq tax_table object
phylo_tree <- phy_tree(genome_tree) 

physeq_genome <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples,phylo_tree)
physeq_genome_clr <- microbiome::transform(physeq_genome, 'clr')

## Wrap working objects

All working objects are wrapped into a single Rdata object to facilitate downstream usage.

```{r wrap_working_objects_gut, warning=FALSE, comments="", message=FALSE, eval=FALSE}
save(sample_metadata, 
     genome_metadata, 
     read_counts, 
     genome_counts, 
     genome_counts_filt, 
     genome_tree,
     physeq_genome,
     phylum_colors,
     file = "data/data.Rdata")
```
git remote set-url origin https://<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>@github.com/alberdilab/squirrels_functional_diversity.git
