---
title: "02_analysis"
output: html_document
date: "2025-02-17"
---
```{r}
load("data/data.Rdata")
singlem <- read.csv("data/singlem.csv",sep=";",header=T)
```
library(tibble)
duplicated_columns <- colnames(genome_gifts)[duplicated(colnames(genome_gifts))]
print(duplicated_columns)
sample_metadata$bases_lost_fastp_percent <- as.numeric(as.character(sample_metadata$bases_lost_fastp_percent))


#data preperation
```{r data_preparation04, message=FALSE}
#Get list of present MAGs
present_MAGs <- genome_counts %>%
  filter(rowSums(.[, -1]) != 0) %>%
  rownames()

#Remove samples with all zeros (no data after filtering)
genome_counts_filt <- genome_counts %>%
  select_if(~!all(. == 0))

#Align distillr annotations with present MAGs and remove all-zero and all-one traits
present_MAGs <- present_MAGs[present_MAGs %in% rownames(genome_gifts)]

genome_gifts <- genome_gifts[, colnames(genome_gifts) != "T050601"]

genome_gifts_filt <- as.data.frame(genome_gifts)[present_MAGs, ] %>%
  select_if(~!all(. == 0)) %>%  # remove all-zero modules
  select_if(~!all(. == 1))  # remove all-one modules

#Align tree with present MAGs
tree_filt <- keep.tip(genome_tree,present_MAGs)

#Filter count table to only contain present MAGs after gifts filtering
genome_counts_filt <- genome_counts[present_MAGs,]

#Calculate sequence fractions for each samples
sequence_fractions <- read_counts %>%
  pivot_longer(-genome, names_to = "sample", values_to = "value") %>%
  group_by(sample) %>%
  dplyr::summarize(mags = sum(value)) %>%
  left_join(sample_metadata, by = join_by(sample == sample))  %>%
  select(sample,mags,metagenomic_bases,host_bases,bases_lost_fastp_percent) %>%
  mutate(mags_bases = mags*146) %>%
  mutate(lowqual_bases = ((metagenomic_bases+host_bases)/(1-bases_lost_fastp_percent))-(metagenomic_bases+host_bases)) %>%
  mutate(unmapped_bases = metagenomic_bases - mags_bases) %>%
  mutate(unmapped_bases = ifelse(unmapped_bases < 0, 0, unmapped_bases)) %>%
  select(sample,mags_bases,unmapped_bases,host_bases,lowqual_bases)
```

#Alpha Diversity
```{r alpha_diversity, cache=TRUE}
#ALPHA DIVERSITY
q0n <- hilldiv2::hilldiv(genome_counts,q=0) %>% c()
q1n <- hilldiv2::hilldiv(genome_counts,q=1) %>% c()
q1p <- hilldiv2::hilldiv(genome_counts,q=1,tree=genome_tree) %>% c()
dist <- hilldiv2::traits2dist(genome_gifts_filt, method="gower")
q1f <- hilldiv2::hilldiv(genome_counts_filt,q=1,dist=dist) %>% c()

```

```{r alpha_summary, warning=FALSE}

# Merge all metrics
alpha <- cbind(sample=colnames(genome_counts),richness=q0n,neutral=round(q1n,3),phylo=round(q1p,3)) %>%
  as.data.frame()
columns <- c("richness","neutral","phylo","func","mapped","total")

# Add amount of sequencing data to the table
alpha <- alpha %>%
  left_join(sequence_fractions, by = join_by(sample == sample)) %>% #add sequencing depth information
  mutate(mapped=round(mags_bases/1000000000,3)) %>% #modify depth to million reads
  mutate(total=round((mags_bases+unmapped_bases+host_bases+lowqual_bases)/1000000000,3)) %>%
  select(sample,richness,neutral,phylo,mapped,total) %>%
  mutate(across(-1, as.numeric))

# Join sample metadata
alpha_div <- alpha %>%
  pivot_longer(-c(sample), names_to = "data", values_to = "value") %>%
  mutate(metric = factor(data, levels = columns)) %>%
  left_join(sample_metadata, by='sample')
 alpha_div %>%
  select(sample, species, metric, value) %>%
  mutate(
    species = factor(species), # Convert species to factor if necessary
    sample = factor(sample, levels = unique(sample[order(species)])) # Reorder sample by species
  ) %>%
  ggplot(aes(x = value, y = sample, fill = species)) +
  geom_bar(stat = 'identity') +
  scale_fill_manual(values = squirrel_colors) +
  facet_wrap(~ metric, scales = "free_x", ncol = 6) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    panel.spacing = unit(0, "lines"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_blank(),
    legend.position = "none"
  )
  
```

#Alpha diversity plots
```{r alpha_sp_plots, fig.dim = c(8, 10)}
neutral.sp <- alpha_div %>%
            filter(metric=="neutral")%>%
            ggboxplot(., x = "species", y = "value", color = "species", fill="white", add="jitter", alpha=0.5) +
                  scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
                  #stat_compare_means() +
                  theme_classic() +
                  labs(y = "Neutral Hill numbers") +
                  theme(
                      legend.position = "none",
                      axis.title.x = element_blank())

phylo.sp <- alpha_div %>%
            filter(metric=="phylo")%>%
            ggboxplot(., x = "species", y = "value", color = "species", fill="white", add="jitter", alpha=0.5) +
                  scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
                  #stat_compare_means() +
                  theme_classic() +
                  labs(y = "Phylogenetic Hill numbers") +
                  theme(
                      legend.position = "none",
                      axis.title.x = element_blank())


ggarrange(neutral.sp, phylo.sp, #+ rremove("x.text"), 
          ncol = 1, nrow = 3)
```

#functional differences
```{r}
# Aggregate bundle-level GIFTs into the compound level
GIFTs_elements <- to.elements(genome_gifts, GIFT_db2)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% rownames(genome_counts), ]
GIFTs_elements_filtered <- as.data.frame(GIFTs_elements_filtered) %>%
  select_if(~ !is.numeric(.) || sum(.) != 0)

elements <- GIFTs_elements_filtered %>%
  as.data.frame()

# Aggregate element-level GIFTs into the function level
GIFTs_functions <- to.functions(GIFTs_elements_filtered, GIFT_db2)
functions <- GIFTs_functions %>%
  as.data.frame()

# Aggregate function-level GIFTs into overall Biosynthesis, Degradation and Structural GIFTs
GIFTs_domains <- to.domains(GIFTs_functions, GIFT_db2)
domains <- GIFTs_domains %>%
  as.data.frame()

# Get community-weighed average GIFTs per sample
GIFTs_elements_community <- to.community(GIFTs_elements_filtered, tss(genome_counts_filt), GIFT_db2)
GIFTs_functions_community <- to.community(GIFTs_functions,tss(genome_counts_filt), GIFT_db2)
GIFTs_domains_community <- to.community(GIFTs_domains,tss(genome_counts_filt), GIFT_db2)
```

```{r}

```

```{r}
GIFTs_functions_community %>%
    as.data.frame() %>%
    rownames_to_column(var="sample") %>%
    pivot_longer(!sample,names_to="trait",values_to="gift") %>%
    left_join(sample_metadata, by = join_by(sample == sample)) %>%
    ggplot(aes(x=trait,y=sample,fill=gift)) +
        geom_tile(colour="white", linewidth=0.2)+
        scale_fill_gradientn(colours=rev(c("#d53e4f", "#f46d43", "#fdae61", "#fee08b", "#e6f598", "#abdda4", "#ddf1da")))+
        facet_grid(species ~ ., scales="free",space="free")
```

```{r}
# Averaged at the function level (each function is weighed equally)
rowMeans(GIFTs_functions)

# Averaged at the domain level (each domain is weighed equally)
rowMeans(GIFTs_domains)
```


#test
```{r}
GIFTs_functions_community_tt <- GIFTs_functions_community %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  separate(sample, into = c("individual", "species"), sep = "\\.") %>%
  pivot_longer(-c(individual, species), names_to = "trait", values_to = "value") %>%
  group_by(trait) %>%
  mutate(model_result = list(lmerTest::lmer(value ~ species + (1 | individual)))) %>%
  ungroup() %>%
  select(trait, model_result) %>%
  unique() %>%
  mutate(estimate = map_dbl(model_result, ~broom.mixed::tidy(.) %>% filter(term == "species") %>% pull(estimate))) %>%
  mutate(p_value = map_dbl(model_result, ~broom.mixed::tidy(.) %>% filter(term == "species") %>% pull(p.value))) %>%
  mutate(p_value_adj = p.adjust(p_value, method = "bonferroni")) %>%
  left_join(GIFT_db %>% select(Code_function, Function) %>% unique(), by = join_by(trait == Code_function)) %>%
  rename(id = trait, trait = Function) %>%
  select(id, trait, estimate, p_value_adj)

GIFTs_functions_community_tt %>% 
  tt() |> 
  style_tt(
    i = which(GIFTs_functions_community_tt$estimate < 0 & GIFTs_functions_community_tt$p_value_adj < 0.05),
    background = "#E5D5B1"
  ) |> 
  style_tt(
    i = which(GIFTs_functions_community_tt$estimate > 0 & GIFTs_functions_community_tt$p_value_adj < 0.05),
    background = "#B7BCCE"
  )

```
#metabolic avegerage plots
```{r}
#prep
GIFTs_domains_community <- GIFTs_domains_community %>%
  as.data.frame() %>%
  left_join(alpha, by="sample") %>%
  mutate(metabolic_function = (Biosynthesis + Degradation) / 2) %>%

GIFTs_domains_diversity <- GIFTs_domains_community %>% 
  as.data.frame() %>% 
  left_join(alpha, by="sample") %>% 
  left_join(sample_metadata, by="sample")%>% 
select(-ends_with(".y")) %>% 
  rename(richness = richness.x, neutral=neutral.x, phylo=phylo.x, mapped=mapped.x, total=total.x) 

 





ggplot(GIFTs_domains_diversity, aes(x=Biosynthesis,y=neutral, color=species)) +
  geom_point() +
  geom_smooth(method="lm")



```


```{r}
#plot: metabolic function and neutral diversity
GIFTs_domains_diversity <- GIFTs_domains_community %>% 
  as.data.frame() %>% 
  left_join(alpha, by="sample") %>% 
  left_join(sample_metadata, by="sample")%>% 
select(-ends_with(".y")) %>% 
  rename(richness = richness.x, neutral=neutral.x, phylo=phylo.x, mapped=mapped.x, total=total.x) 

ggplot(GIFTs_domains_diversity, aes(x=metabolic_function,y=neutral, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
  geom_point()+
  geom_smooth(method = "lm")
```
```{r}
#plot: metabolic function and phylogenetic diversity
ggplot(GIFTs_domains_diversity, aes(x=metabolic_function,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
  geom_point()+
  geom_smooth(method = "lm")
```

```{r}
#plot: sciurus carolinensis, linear model
GIFTs_domains_diversity <- GIFTs_domains_diversity %>% 
     filter(species=="Sciurus carolinensis")

ggplot(GIFTs_domains_diversity, aes(x=metabolic_function,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
  geom_point()+
  geom_smooth(method = "lm")

cor.test(GIFTs_domains_diversity$metabolic_function,GIFTs_domains_diversity$phylo, method = "pearson")
```
```{r}
#plot: sciurus vulgrais, linear model
GIFTs_domains_diversity <- GIFTs_domains_diversity %>% 
     filter(species=="Sciurus vulgaris")

ggplot(GIFTs_domains_diversity, aes(x=metabolic_function,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
  geom_point()+
  geom_smooth(method = "lm")

cor.test(GIFTs_domains_diversity$metabolic_function,GIFTs_domains_diversity$phylo, method = "pearson")
```
```{r}
#Biosynthesis
ggplot(GIFTs_domains_diversity, aes(x=Biosynthesis,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
  geom_point()+
  geom_smooth(method = "lm")
```
```{r}
#Degradation
ggplot(GIFTs_domains_diversity, aes(x=Degradation,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
  geom_point()+
  geom_smooth(method = "lm")
```


# Individual metabolic traits
```{r}
GIFTs_functions_community_collected <- GIFTs_functions_community_tt %>% 
  as.data.frame() %>% 
  left_join(GIFTs_domains_diversity, by="sample")

#B01
ggplot(GIFTs_functions_community_collected, aes(x=B01,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
geom_point()+
  geom_smooth(method="lm")

```
```{r}
#B02
ggplot(GIFTs_functions_community_collected, aes(x=B02,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
geom_point()+
  geom_smooth(method="lm")
```
```{r}
#B03
ggplot(GIFTs_functions_community_collected, aes(x=B03,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
geom_point()+
  geom_smooth(method="lm")
```
```{r}
#B04
ggplot(GIFTs_functions_community_collected, aes(x=B04,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
geom_point()+
  geom_smooth(method="lm")
```
```{r}
#B06
ggplot(GIFTs_functions_community_collected, aes(x=B06,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
geom_point()+
  geom_smooth(method="lm")
```
```{r}
#B07
ggplot(GIFTs_functions_community_collected, aes(x=B07,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
geom_point()+
  geom_smooth(method="lm")
```
```{r}
#B08
ggplot(GIFTs_functions_community_collected, aes(x=B08,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
geom_point()+
  geom_smooth(method="lm")
```
```{r}
#B09
ggplot(GIFTs_functions_community_collected, aes(x=B09,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
geom_point()+
  geom_smooth(method="lm")
```
```{r}
#B10
ggplot(GIFTs_functions_community_collected, aes(x=B10,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
geom_point()+
  geom_smooth(method="lm")
```
```{r}
#B11
ggplot(GIFTs_functions_community_collected, aes(x=B11,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
geom_point()+
  geom_smooth(method="lm")
```

```{r}
#D01
ggplot(GIFTs_functions_community_collected, aes(x=D01,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
geom_point()+
  geom_smooth(method="lm")
```
```{r}
#D02
ggplot(GIFTs_functions_community_collected, aes(x=D02,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
geom_point()+
  geom_smooth(method="lm")
```
```{r}
#D03
ggplot(GIFTs_functions_community_collected, aes(x=D03,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
geom_point()+
  geom_smooth(method="lm")
```
```{r}
#D05
ggplot(GIFTs_functions_community_collected, aes(x=D05,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
geom_point()+
  geom_smooth(method="lm")
```
```{r}
#D06
ggplot(GIFTs_functions_community_collected, aes(x=D06,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
geom_point()+
  geom_smooth(method="lm")
```
```{r}
#D07
ggplot(GIFTs_functions_community_collected, aes(x=D07,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
geom_point()+
  geom_smooth(method="lm")
```
```{r}
#D08
ggplot(GIFTs_functions_community_collected, aes(x=D08,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
geom_point()+
  geom_smooth(method="lm")
```
```{r}
#D09
ggplot(GIFTs_functions_community_collected, aes(x=D09,y=phylo, color=species))+scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
geom_point()+
  geom_smooth(method="lm")
```

```{r}
#correlations of traits

# Example: Assuming df contains phylo and the selected traits, and 'species' column exists
traits <- c("B01", "B02", "B03", "B04", "B06", "B07", "B08", "B09", 
            "B10", "B11", "D01", "D02", "D03", "D05", "D06", "D07", "D08", "D09")

# Create an empty dataframe to store results
correlation_results <- data.frame(Trait = rep(traits, each = length(unique(GIFTs_functions_community_collected$species))),
                                  Species = rep(unique(GIFTs_functions_community_collected$species), times = length(traits)), 
                                  Correlation = NA, P_value = NA, Significance = NA)

# Loop through each trait
for (i in seq_along(traits)) {
  trait <- traits[i]
  
  # Loop through each species
  for (species_name in unique(GIFTs_functions_community_collected$species)) {
    # Subset the data for the current species
    species_data <- GIFTs_functions_community_collected %>% filter(species == species_name)
    
    # Compute the correlation between phylo and the current trait for the current species
    test <- cor.test(species_data$phylo, species_data[[trait]], method = "pearson")  # Change to "spearman" if needed
    
    # Store the correlation, p-value, and significance symbol
    correlation_results$Correlation[correlation_results$Trait == trait & correlation_results$Species == species_name] <- test$estimate
    correlation_results$P_value[correlation_results$Trait == trait & correlation_results$Species == species_name] <- test$p.value
    
    # Add a significance symbol or label based on the p-value
    correlation_results$Significance[correlation_results$Trait == trait & correlation_results$Species == species_name] <- 
      ifelse(test$p.value < 0.05, "*", "")
  }
}

# Print the correlation table with significance symbol
print(correlation_results)



```


```{r}
#genome size and metabolic capacity plot

CWA_genome_size <- genome_counts_by_host %>%
  group_by(sample) %>%  # Group by sample to calculate per sample
  summarise(cwa_genome_size = sum(corrected_size * count)) %>%  # Weighted average
  ungroup()

merged_data <- GIFTs_functions_community_collected %>% 
  as.data.frame() %>% 
  left_join(CWA_genome_size, by="sample")
  
  ggplot(merged_data, aes(x=metabolic_function, y=cwa_genome_size, color=species)) + 
  scale_color_manual(values=squirrel_colors) +
  scale_fill_manual(values=paste0(squirrel_colors)) +
  geom_point() +
  geom_smooth(method="lm")


```

```{r}
#correlation tests
merged_data <- merged_data%>%
  filter(species=="Sciurus vulgaris")

cor.test(merged_data$metabolic_function, merged_data$cwa_genome_size, method = "pearson")

merged_data <- merged_data%>%
  filter(species=="Sciurus carolinensis")

cor.test(merged_data$metabolic_function, merged_data$cwa_genome_size, method = "pearson")


```


